---
title: "Multivariate Data Analysis Project"
author: "Adrián Cerezuela and Ramon Venutra"
date: "5/22/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("~/Desktop/GCED/obsidian/GCED/Q4/AD/Multivariate_Analysis_Project/")
```

```{r}
library(ggplot2)
library(MASS)
library(GGally)
#library(tidyverse)
library(factoextra)
library(FactoMineR)
library(cabootcrs)
library(caret)
library(cluster)
library(kmed)
library(fpc)
#library(HSAUR)
library(biotools)
library(klaR)
```


### Dataset preprocessing

Firstly, we read the dataset.
```{r}
data_acb <- read.csv("data_preprocessed.csv")
head(data_acb)
```

With the help of the $\texttt{str()}$ function, the correct variable type will be assigned to each row.
```{r}
str(data_acb)
```

```{r}
data_acb[,4] <- as.numeric(data_acb[,4])

str(data_acb)
```


### PRINCIPAL COMPONENTS ANALYSIS (PCA)

```{r}
numerical_vars <- scale(data_acb[, c(3,4,7:21)])
```

```{r}
pca_acb = prcomp(numerical_vars)
```

```{r}
fviz_eig(pca_acb, addlabels = TRUE) # explained variability plot
```

```{r}
biplot <- fviz_pca_biplot(pca_acb, repel = F, axes = c(2,3), col.var = "black",
                          habillage = data_acb$Performance)
# Needs to be reflected on both axis taking into account the contributions
(reversed_biplot <- biplot + scale_x_reverse() + scale_y_reverse())
```

```{r}
# Plot the PCA graph of variables
var_plot <- fviz_pca_var(pca_acb,
                         col.var = "contrib", # color variables by contribution
                         gradient.cols = c("blue", "red"), # specify color gradient
                         repel = TRUE # avoid overlapping labels
)

# Needs to be reflected too
(reflected_var_plot <- var_plot + scale_x_reverse() + scale_y_reverse())
```

```{r}
(loadings <- get_pca_var(pca_acb)$coord)
```

```{r}
(contributions <- apply(pca_acb$rotation^2, 2, function(x) x/sum(x))*100)
```
```{r}
(col_sums <- colSums(contributions))
```



### MULTIDIMENSIONAL SCALING (MDS)

```{r}
dist_matrix <- dist(numerical_vars, method = "euclidean")

pos <- data_acb[,2]
perf <- data_acb[,22]
new_data = as.data.frame(numerical_vars)
new_data$pos = as.factor(pos)
new_data$perf = as.factor(perf)
gower_dist <- daisy(new_data, metric = "gower")
```

```{r}
mds_euc <- cmdscale(dist_matrix, eig=T, k = 2)
mds_gow <- cmdscale(gower_dist, eig=T, k = 2)

# Colors for each performance
cluster_colors_perf <- c("red", "green", "blue")
```


```{r}
#Cluster Performance
mds_data1 <- data.frame(x = mds_euc$points[,1], y = mds_euc$points[,2], cluster = perf)
mds_data2 <- data.frame(x = mds_gow$points[,1], y = mds_gow$points[,2], cluster = perf)

par(mfrow=c(1,2))

ggplot(mds_data1, aes(x = x, y = y, color = as.factor(cluster))) +
  geom_point(size = 3) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot (Euclidean distance Matrix) with Clusters")

ggplot(mds_data2, aes(x = x, y = y, color = as.factor(cluster))) +
  geom_point(size = 3) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot (Gower Distance Matrix) with Clusters")

par(mfrow=c(1,1))
```

```{r}
# Label PPG
ggplot(mds_data2, aes(x = x, y = y, label=data_acb[,7], color = as.factor(cluster))) +
  geom_point(size = 2) +
  geom_text(vjust=-0.5) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot with Clusters and PPG as labels")
```
```{r}
# Label X3PT_Per
ggplot(mds_data2, aes(x = x, y = y, label=data_acb[,10], color = as.factor(cluster))) +
  geom_point(size = 2) +
  geom_text(vjust=-0.5) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot with Clusters and X3PT_Per as labels")
```
```{r}
# Label X2PT_Per
ggplot(mds_data2, aes(x = x, y = y, label=data_acb[,11], color = as.factor(cluster))) +
  geom_point(size = 2) +
  geom_text(vjust=-0.5) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot with Clusters and X2PT_Per as labels")
```
```{r}
# Label TR
ggplot(mds_data2, aes(x = x, y = y, label=data_acb[,13], color = as.factor(cluster))) +
  geom_point(size = 2) +
  geom_text(vjust=-0.5) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot with Clusters and TR as labels")
```
```{r}
# Label BLKF
ggplot(mds_data2, aes(x = x, y = y, label=data_acb[,17], color = as.factor(cluster))) +
  geom_point(size = 2) +
  geom_text(vjust=-0.5) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot with Clusters and BLKF as labels")
```
```{r}
# Label AST
ggplot(mds_data2, aes(x = x, y = y, label=data_acb[,14], color = as.factor(cluster))) +
  geom_point(size = 2) +
  geom_text(vjust=-0.5) +
  scale_color_manual(values = cluster_colors_perf) +
  xlab("MDS Coordinate 1") +
  ylab("MDS Coordinate 2") +
  ggtitle("MDS Plot with Clusters and AST as labels")
```

## MCA

```{r}
## Indicator Matrix
# ind_matrix <- tab.disjonctif(data_acb[, categorical_vars])
```

Correct data format:
```{r}
categorical_vars <- c("pos", "license", "club", "Performance")
data_acb$pos <- as.factor(data_acb$pos)
data_acb$license <- as.factor(data_acb$license)
data_acb$club <- as.factor(data_acb$club)
data_acb$Performance <- as.factor(data_acb$Performance)
```

MCA on categorical variables
```{r}
# MCA Application on Indicator Matrix
res.mca <- MCA(data_acb[, categorical_vars]) # Cloud of individuals, variables
# res.mca <- MCA(data_acb[, categorical_vars],method="Burt") # Using Burt Matrix instead

# summary(res.mca)

### Eigenvalues
# res.mca$eig

## Average eigenvalue
```

Eigenvalues
```{r}
sum(res.mca$eig[,1])/12
sum(res.mca$eig[,2]) / length(res.mca$eig[,2]) # % variance
barplot(res.mca$eig[,2],main="Eigenvalues", names.arg=1:nrow(res.mca$eig))
```

```{r}
## Coordinates of categories
res.mca$var$coord

## Description of Dimensions
dimdesc(res.mca)
```

```{r}
rownames(res.mca$var$coord) # All the categories
```

Clouds of individuals
```{r}
### Clouds (besides from the ones generated from MCA())###
## Colored cloud of Categories
plot(res.mca, invisible = "ind",
     col.var = c(rep("black", length(unique(data_acb$pos))),
                 rep("red", length(unique(data_acb$license))),
                 rep("blue", length(unique(data_acb$club))),
                 rep("purple", length(unique(data_acb$Performance)))),
     title = "Graph of the active categories")


## Cloud of Individuals
plot.MCA(res.mca,choix="ind",label="none")
plot.MCA(res.mca,choix="ind",label="ind",invisible="var",habillage = "pos")
plot.MCA(res.mca,choix="ind",label="ind",invisible="var",habillage = "license")
plot.MCA(res.mca,choix="ind",label="ind",invisible="var",habillage = "club")
plot.MCA(res.mca,choix="ind",label="ind",invisible="var",habillage = "Performance")
```

```{r}
str(data_acb)
```


MCA on data_acb using numerical variables as supplementary
```{r}
res.mca <- MCA(data_acb[, -c(1,21)], quanti.sup=c(2:3,6:19))
```

Eigenvalues using supplementary variables:
```{r}
sum(res.mca$eig[,1])/12
sum(res.mca$eig[,2]) / length(res.mca$eig[,2]) # % variance
barplot(res.mca$eig[,2],main="Eigenvalues", names.arg=1:nrow(res.mca$eig))
```


### CLUSTER ANALYSIS

We will perform clusters on the diferent teams we have, in order to asses which teams have better performers in their roster.

```{r}
data <- data_acb[, c(6,3,4,7:21)]
averaged_data <- aggregate(. ~ club, data = data, FUN = mean)

rownames(averaged_data)<-averaged_data[,1]
data<-averaged_data[,-1]
```

Arbitrarily, we could think of at least k=3 clusters having in account a separation between an encompassed team performance

```{r}
k <- 3
k3 <- kmeans(data, centers=k)
aggregate(data,by=list(k3$cluster),FUN=mean)
```


```{r}
fviz_cluster(k3, data, geom="point")
```

We can observe how data is not separed at all, so we perform an elbow plot and Pseudo-F Index plot in order to decide the optimal number of clusters.

```{r}
# Elbow plot
aux<-c()
for (i in 2:10){
  k<-kmeans(data,centers=i,nstart=25)
  # centers: number of clusters to create in the k-means algorithm
  # nstart: number of random initializations to perform for the k-means algorithm. This helps find a good solution by avoiding local optima.
  aux[i-1]<-k$tot.withinss
}
### Pseudo-F Index
aux2<-c()
for (i in 2:10){
  k<-kmeans(data,centers=i,nstart=25)
  aux2[i-1]<-((k$betweenss)*(nrow(data)-i))/((k$tot.withinss)*(i-1))
}
par(mfrow=c(1,2))
plot(aux, xlab="Number of Clusters", ylab="TWSS", type="l", main="TWSS vs. number of clusters")
plot(aux2, xlab="Number of Clusters", ylab="Pseudo-F", type="l", main="Pseudo F Index")
par(mfrow=c(1,1))
```
From these two plots one can intuit that the optimal number of clusters are k=7. Let's see how k-means perform for 7 clusters and the difference with the previous one.

```{r}
k <- 7
k7 <- kmeans(data, centers=k)
fviz_cluster(k7, data, geom="point")
```

We can see how k=7 clusters does not bring any improvement from the first k=3 performed, so we are taking the first one. Through the principal components on our data (without aggregating), we can see how the clusters are separated.

```{r}
clusters <- k3$cluster
pca_acb <- prcomp(data)
pc_scores <- pca_acb$x

(ggplot(data = as.data.frame(pc_scores), aes(x = PC1, y = PC2, color = as.factor(clusters))) +
  geom_point() +
  geom_text(aes(label = rownames(averaged_data))) +
  geom_polygon(data = as.data.frame(pc_scores),
               aes(x = PC1, y = PC2, fill = as.factor(clusters), group = as.factor(clusters)),
               alpha = 0.2) +
  scale_color_discrete(name = "Cluster") +
  scale_fill_discrete(name = "Cluster") +
  theme_bw())
```
Hierarchical Clustering

```{r}
d <- dist(data, method = "manhattan")
```

```{r}
# Complete Linkage
fit <- hclust(d, method="complete") 
plot(fit, label=row.names(averaged_data), main="Dendrogram of Complete Linkage")
rect.hclust(fit, k=3, border="blue")
```

```{r}
# Average Linkage
fit <- hclust(d, method="average") 
plot(fit, label=row.names(averaged_data), main="Dendrogram of Average Linkage")
rect.hclust(fit, k=3, border="blue")
```
```{r}
cl = cutree(fit, k = 3)
fviz_cluster(list(data = data, cluster = cl), geom="point")
```

### LDA/QDA

```{r}
lda_data <- data_acb[, c(3,4,7:22)]
lda_data <- lda_data[,-17]
```

```{r}
### Assumption 1: Normality ####
## Checking histogram, qqplots and shapiro Wilk's test results
par(mfrow = c(3,4)) 
for (i in c(1:16)){ 
  print(colnames(lda_data)[i])
  print(shapiro.test(lda_data[,i])$p.value)
  qqnorm(lda_data[,i],main=colnames(lda_data)[i])
  hist(lda_data[,i], main=colnames(lda_data)[i])
}
```
```{r}
boxM(lda_data[,1:16],lda_data$Performance)
```
As percentages are not useful to predict the performance index rating of a player, and therefore not useful to predict our target variable, are not included in our model. An important reason for not doing it is also that observations are far from following a normal distribution.

```{r}
(acblda<-lda(Performance~height+age+PPG+FGM+FGA+TR+AST+STL+TOV+
              BLKF+BLKA+FC+FD, data=lda_data))
acblda$prior
```
```{r}
acbpred<-predict(acblda)
(confusionlda<-table(acbpred$class,data_acb$Performance))
```
```{r}
# CCR
(classrate<-sum(diag(confusionlda))/sum(confusionlda))
```
```{r}
# CCR among groups
(diag(prop.table(confusionlda, 1)))
```

```{r}
(acbqda<-qda(Performance~height+age+PPG+FGM+FGA+TR+AST+STL+TOV+
              BLKF+BLKA+FC+FD, data=lda_data))
```
```{r}
qda.class <- predict (acbqda)
(confusionqda<-table(qda.class$class,data_acb$Performance))
```
```{r}
(classrate<-sum(diag(confusionqda))/sum(confusionqda))
```

```{r}
# Plot the results of our classification prediction for LDA
grouping_var <- lda_data$Performance

plot_data <- data.frame(LDA = as.factor(acbpred$class), Performance = as.factor(grouping_var))

# Plot using ggplot2
ggplot(plot_data, aes(x = Performance, fill = LDA)) +
  geom_bar(position = "fill") +
  labs(x = "Performance", y = "Proportion") +
  scale_fill_manual(values = c("red", "green", "blue"))
```
```{r}
# Plot the results of our classification prediction for QDA
plot_data <- data.frame(QDA = as.factor(qda.class$class), Performance = as.factor(grouping_var))

# Plot using ggplot2
ggplot(plot_data, aes(x = Performance, fill = QDA)) +
  geom_bar(position = "fill") +
  labs(x = "Performance", y = "Proportion") +
  scale_fill_manual(values = c("red", "green", "blue"))
```

```{r}
# Q=(n-ñk)^2/n(k-1)
## compared with chi^2 value of (1,0.05)
nyq<-sum(diag(confusionqda))
nc<-nrow(lda_data)
k<-3
(Qcar<-((nc-nyq*k)^2)/(nc*(k-1)))
```
Too high, difference in cov matrices.


```{r}
### Naive Bayes Classification ##
#acbnb<-naiveBayes(Performance~height+age+PPG+FGM+FGA+TR+AST+STL+TOV+
 #             BLKF+BLKA+FC+FD, data=lda_data)

#nb.class <- predict (acbnb , lda_data)
#(confusionnb <- table(nb.class, data_acb$Performance))
```
```{r}
#(classrate<-sum(diag(confusionnb))/sum(confusionnb))
```

```{r}
# Stepwise Classification
lda_data <- lda_data[,-c(6,7,8)]
acbstep<-stepclass(lda_data[,1:13],lda_data$Performance,method="qda",direction="backward", criterion="CR") 

acbstep$process
acbstep$model
acbstep$result.pm
```






























